---
title: "Practical Machine Learning - A prediction model for weight lifting exercises"
author: "Ian Dobbs"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
# knitr options
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, fig.cap = TRUE, fig.align = "center",
                      fig.path="figures/", options(scipen=999))
knitr::opts_current$get('label')
```

```{r captioner, message=FALSE, echo=FALSE, results='hide'}
# use captioner to add figure number and caption
library(captioner)
fig_nums <- captioner()
fig_nums("figa", "")
fig_nums("figb", "")
fig_nums("figc", "")
```

# Executive Summary

This purpose of this report is to predict how well 6 participants aged between 20-28 years perform a Unilateral Dumbbell Biceps curl. Using data from inertial measurement units (IMU) in the users glove, armband, lumbar belt and dumbbell participants were asked to perform one set of 10 repetitions using correct and incorrect weight lifting technique (producing 5 'classes' corresponding to correctly specified execution and 4 common mistakes). The data was explored to produce a prediction model capable of predicting how well the exercise will be performed based on a separate testing dataset.

_Note that the `echo = FALSE` parameter has been added to the code chunks to prevent printing of the R code, all of which can be found in the appendix._

## Summary of the data
```{r wle}
# load the dataset and display the structure
pml_training <- read.csv("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv")
pml_testing <- read.csv("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv")
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
